Case Study: ChIP-seq Differential Peak Calling (csaw)

### 2018-08-29 datasets/ChIP-seq/ChIPseq-CBP-csaw.Rmd
Fix typo in ChIPseq; adjust pvalue histograms
Update ChiPseq csaw analyses
Add mean coverage (across all samples) as an informative covariate.
In addition, update call to csaw::normOffsets to reflect changes in
current release version that change the default return object to a
SE.

```{r, wkspace-setup, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(SummarizedBenchmark)
library(BiocParallel)
library(rtracklayer)
library(Rsamtools)
library(Rsubread)
library(csaw)
library(edgeR)

@@

## intermediary files we create below
count_file <- file.path(resdir, "cbp-csaw-counts.rds")
filtered_file <- file.path(resdir, "cbp-csaw-counts-filtered.rds")
normfacs_file <- file.path(resdir, "cbp-csaw-counts-normfacs.rds")
result_file <- file.path(resdir, "cbp-csaw-results.rds")
bench_file <- file.path(sbdir, "cbp-csaw-benchmark.rds")
bench_file_uninf <- file.path(sbdir, "cbp-csaw-uninf-benchmark.rds")

@@

    ## calculate normalizing offsets based on larger windows
    normfacs <- normOffsets(bins)

@@

    ## save results
    res_ranges <- merged$region
    elementMetadata(res_ranges) <-
        data.frame(tab_comb,
                   best_pos = mid(ranges(rowRanges(filtered_cnts[tab_best$best]))),
                   best_logFC = tab_best$logFC)
    saveRDS(res_ranges, file = result_file)
}

@@

```{r, nwin-diag-hist, fig.width=10, fig.height=3.2}
strat_hist(res_df, pvalue = "PValue", covariate = "nWindows", maxy = 35, numQ = 3)
```

@@

```{r, width-diag-hist, fig.width=10, fig.height=3.2}
strat_hist(res_df, pvalue = "PValue", covariate = "width", maxy = 35, numQ = 3)
```

@@

```{r, rand-diag-hist, fig.width=10, fig.height=3.2}
strat_hist(res_df, pvalue = "PValue", covariate = "uninf_covar", maxy = 35, numQ = 3)
```

@@

We use the common `BenchDesign` with the set of multiple testing correction
methods already included. We use the width of the regions as the independent
covariate.

@@

@@

@@

```{r, chipseq-overlap-rand}
plotFDRMethodsOverlap(sbU, alpha = 0.05, nsets = ncol(sb),
                      order.by = "freq", decreasing = TRUE,
                      supplementary = FALSE)
```

@@

```{r}
plotMethodRanks(c(benchfile, benchfile_uninf), 
                colLabels = c("Region width", "Random"), 
                alpha = 0.10, xlab = "Covariate", 
                excludeMethods = NULL)
```

>>>>

```{r, wkspace-setup, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(SummarizedBenchmark)
library(BiocParallel)
library(rtracklayer)
library(Rsamtools)
library(Rsubread)
library(csaw)
library(edgeR)
library(GenomicAlignments)

@@

## intermediary files we create below
count_file <- file.path(resdir, "cbp-csaw-counts.rds")
filtered_file <- file.path(resdir, "cbp-csaw-counts-filtered.rds")
normfacs_file <- file.path(resdir, "cbp-csaw-counts-normfacs.rds")
result_file <- file.path(resdir, "cbp-csaw-results.rds")
bench_file <- file.path(sbdir, "cbp-csaw-benchmark.rds")
bench_file_cov <- file.path(sbdir, "cbp-csaw-cov-benchmark.rds")
bench_file_uninf <- file.path(sbdir, "cbp-csaw-uninf-benchmark.rds")

@@

    ## calculate normalizing offsets based on larger windows
    normfacs <- normOffsets(bins, se.out = FALSE)

@@

    ## save results
    res_ranges <- merged$region
    elementMetadata(res_ranges) <-
        data.frame(tab_comb,
                   best_pos = mid(ranges(rowRanges(filtered_cnts[tab_best$best]))),
                   best_logFC = tab_best$logFC)

    ## get overall mean counts for each window
    merged_cnts <- summarizeOverlaps(res_ranges, BamFileList(meta$sortedfile))
    res_ranges$meancnt <- rowMeans(assays(merged_cnts)$counts)

    saveRDS(res_ranges, file = result_file)
}

@@

```{r, nwin-diag-hist, fig.width=10, fig.height=3.2}
strat_hist(res_df, pvalue = "PValue", covariate = "nWindows", maxy = 15)
```

@@

```{r, width-diag-hist, fig.width=10, fig.height=3.2}
strat_hist(res_df, pvalue = "PValue", covariate = "width", maxy = 15)
```

### Mean Coverage

```{r, cov-diag-scatter, fig.width=4.5, fig.height=3.5}
rank_scatter(res_df, pvalue = "PValue", covariate = "meancnt") +
    ggtitle("Mean coverage as independent covariate") +
    xlab("Mean coverage")
```

```{r, cov-diag-hist, fig.width=10, fig.height=3.2}
strat_hist(res_df, pvalue = "PValue", covariate = "meancnt", maxy = 15)
```

@@

```{r, rand-diag-hist, fig.width=10, fig.height=3.2}
strat_hist(res_df, pvalue = "PValue", covariate = "uninf_covar", maxy = 15)
```

@@

We use the common `BenchDesign` with the set of multiple testing correction
methods already included. We investigate both the width of the regions and their
mean coverage as the independent covariate. We won't assess the number of windows,
since this is very tightly correlated with the width of the region.

```{r}
cor(res_df[,c("width", "nWindows", "meancnt")])
```

First, we'll use the region width covariate.

@@

Next, we'll use the mean coverage covariate.

```{r}
if (!file.exists(bench_file_cov)) {
    res_df$ind_covariate <- res_df$meancnt
    res_df$pval <- res_df$PValue

    bd <- initializeBenchDesign()
    sbC <- buildBench(bd, data = res_df, ftCols = "ind_covariate")
    saveRDS(sbC, file = bench_file_cov)
} else {
    sbC <- readRDS(bench_file_cov)
}
```

@@

### Mean Coverage

We'll do the same for the mean coverage covariate.

```{r, chipseq-metrics-cov}
assayNames(sbC) <- "qvalue"
sbC <- addDefaultMetrics(sbC)
```

```{r, chipseq-scatter-cov, results="hide"}
rejections_scatter(sbC, supplementary = FALSE)
rejection_scatter_bins(sbC, covariate = "ind_covariate",
                       bins = 4, supplementary = FALSE)
```

```{r, chipseq-overlap-cov}
plotFDRMethodsOverlap(sbC, alpha = 0.05, nsets = ncol(sbC),
                      order.by = "freq", decreasing = TRUE,
                      supplementary = FALSE)
```

```{r, rnaseq-lineplot-cov}
covariateLinePlot(sbC, alpha = 0.05, covname = "ind_covariate")
```

@@

```{r, chipseq-overlap-rand}
plotFDRMethodsOverlap(sbU, alpha = 0.05, nsets = ncol(sbU),
                      order.by = "freq", decreasing = TRUE,
                      supplementary = FALSE)
```

@@

```{r}
plotMethodRanks(c(bench_file, bench_file_cov, bench_file_uninf), 
                colLabels = c("Region width", "Mean Coverage", "Random"), 
                alpha = 0.10, xlab = "Covariate", 
                excludeMethods = NULL)
```

### 2018-08-29 datasets/ChIP-seq/ChIPseq-h3k4me3-csaw.Rmd
Update ChiPseq csaw analyses
Add mean coverage (across all samples) as an informative covariate.
In addition, update call to csaw::normOffsets to reflect changes in
current release version that change the default return object to a
SE.

## intermediary files we create below
count_file <- file.path(resdir, "h3k4me3-csaw-counts.rds")
filtered_file <- file.path(resdir, "h3k4me3-csaw-counts-filtered.rds")
result_file <- file.path(resdir, "h3k4me3-csaw-results.rds")
bench_file <- file.path(sbdir, "h3k4me3-csaw-benchmark.rds")
bench_file_uninf <- file.path(sbdir, "h3k4me3-csaw-uninf-benchmark.rds")

@@

```{r test, results='hide', message=FALSE, warning=FALSE}
if (file.exists(result_file)) {
    res_ranges <- readRDS(result_file)
} else {
    ## normalize for library-specific trended biases
    offsets <- normOffsets(filtered_cnts, type = "loess")

@@

    ## run quasi-likelihood F-test
    y <- asDGEList(filtered_cnts)
    y$offset <- offsets
    y <- estimateDisp(y, design)
    fit <- glmQLFit(y, design, robust = TRUE)
    res <- glmQLFTest(fit)

@@

@@

```{r, nwin-diag-hist, fig.width=10, fig.height=3.2}
strat_hist(res_df, pvalue = "PValue", covariate = "nWindows", maxy = 35, numQ = 3)
```

@@

```{r, width-diag-hist, fig.width=10, fig.height=3.2}
strat_hist(res_df, pvalue = "PValue", covariate = "width", maxy = 35, numQ = 3)
```

@@

```{r, rand-diag-hist, fig.width=10, fig.height=3.2}
strat_hist(res_df, pvalue = "PValue", covariate = "uninf_covar", maxy = 35, numQ = 3)
```

@@

@@

```{r}
plotMethodRanks(c(bench_file, bench_file_uninf), 
                colLabels = c("Region width", "Random"), 
                alpha = 0.10, xlab = "Covariate", 
                excludeMethods = NULL)
```

>>>>

## intermediary files we create below
count_file <- file.path(resdir, "h3k4me3-csaw-counts.rds")
filtered_file <- file.path(resdir, "h3k4me3-csaw-counts-filtered.rds")
result_file <- file.path(resdir, "h3k4me3-csaw-results.rds")
bench_file <- file.path(sbdir, "h3k4me3-csaw-benchmark.rds")
bench_file_cov <- file.path(sbdir, "h3k4me3-csaw-cov-benchmark.rds")
bench_file_uninf <- file.path(sbdir, "h3k4me3-csaw-uninf-benchmark.rds")

@@

```{r test, results='hide', message=FALSE, warning=FALSE}
if (file.exists(result_file)) {
    res_ranges <- readRDS(result_file)
} else {
    ## normalize for library-specific trended biases
    filtered_cnts <- normOffsets(filtered_cnts, type = "loess")

@@

    ## run quasi-likelihood F-test
    y <- asDGEList(filtered_cnts)
    y <- estimateDisp(y, design)
    fit <- glmQLFit(y, design, robust = TRUE)
    res <- glmQLFTest(fit)

@@

    ## get overall mean counts for each window
    merged_cnts <- summarizeOverlaps(res_ranges, BamFileList(meta$sortedfile))
    res_ranges$meancnt <- rowMeans(assays(merged_cnts)$counts)

@@

```{r, nwin-diag-hist, fig.width=10, fig.height=3.2}
strat_hist(res_df, pvalue = "PValue", covariate = "nWindows", maxy = 35)
```

@@

```{r, width-diag-hist, fig.width=10, fig.height=3.2}
strat_hist(res_df, pvalue = "PValue", covariate = "width", maxy = 35)
```

### Mean Coverage

```{r, cov-diag-scatter, fig.width=4.5, fig.height=3.5}
rank_scatter(res_df, pvalue = "PValue", covariate = "meancnt") +
    ggtitle("Mean coverage as independent covariate") +
    xlab("Mean coverage")
```

```{r, cov-diag-hist, fig.width=10, fig.height=3.2}
strat_hist(res_df, pvalue = "PValue", covariate = "meancnt", maxy = 35)
```

@@

```{r, rand-diag-hist, fig.width=10, fig.height=3.2}
strat_hist(res_df, pvalue = "PValue", covariate = "uninf_covar", maxy = 35)
```

@@

### Mean coverage

We'll do the same for the mean coverage covariate.

```{r, chipseq-metrics-cov}
assayNames(sbC) <- "qvalue"
sbC <- addDefaultMetrics(sbC)
```

```{r, chipseq-scatter-cov, results="hide"}
rejections_scatter(sbC, supplementary = FALSE)
rejection_scatter_bins(sbC, covariate = "ind_covariate",
                       bins = 4, supplementary = FALSE)
```

```{r, chipseq-overlap-cov}
plotFDRMethodsOverlap(sbC, alpha = 0.05, nsets = ncol(sbC),
                      order.by = "freq", decreasing = TRUE,
                      supplementary = FALSE)
```

```{r, rnaseq-lineplots-cov}
covariateLinePlot(sbC, alpha = 0.05, covname = "ind_covariate")
```

@@

```{r}
plotMethodRanks(c(bench_file, bench_file_cov, bench_file_uninf), 
                colLabels = c("Region width", "Mean Coverage", "Random"), 
                alpha = 0.10, xlab = "Covariate", 
                excludeMethods = NULL)
```
